Pyro -Velocity: Probabilistic RNA Velocity inference from single-cell data

### 2024-08-21 reproducibility/manuscript/v2.tex
chore(manuscript): update v2 diff file

\usepackage{graphicx}%
\usepackage{multirow}%
\usepackage{amsmath,amssymb,amsfonts}%
\usepackage{amsthm}%
\usepackage{mathrsfs}%
\usepackage[title]{appendix}%
\usepackage{xcolor}%
\usepackage{textcomp}%
\usepackage{manyfoot}%
\usepackage{booktabs}%
\usepackage{algorithm}%
\usepackage{algorithmicx}%
\usepackage{algpseudocode}%
\usepackage{listings}%

@@

\raggedbottom
\usepackage{nameref}
\usepackage{placeins}
\allowdisplaybreaks
\makeatletter
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{plain}
\@ifundefined{c@chapter}{\newfloat{suppfig}{h}{loextfig}}{\newfloat{suppfig}{h}{loextfig}[chapter]}
\floatname{suppfig}{Extended Data Fig.}
\newcommand*\listofsuppfigs{\listof{suppfig}{List of Extended Data Figures}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

@@

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Pyro -Velocity: Probabilistic RNA Velocity inference from single-cell data},
  pdfkeywords={RNA velocity, single-cell RNA sequencing, probabilistic
generative modeling, cell fate, developmental trajectory inference},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{document}
\maketitle
\newpage{}

\renewcommand*\contentsname{Contents}
{
\hypersetup{linkcolor=}
\setcounter{tocdepth}{2}
\tableofcontents
}

@@

We next used Pyro -Velocity to unbiasedly uncover potential cell fate
determinant markers genes within the E15.5 mouse pancreas. To this end,
we first selected the top 300 genes with the best goodness of fit
(negative mean absolute errors across cells). Next, we ranked them based
on the positive (or negative) correlation between inferred shared time
and corresponding denoised spliced expression per cell
(Fig.~\ref{fig-model-results}d, Methods). This procedure successfully
recovered key developmental genes that are activated (or repressed)
during endocrinogenesis (Fig.~\ref{fig-model-results}d-e,
Extended Data Fig.~\ref{extfig-model-results}). Among the top positively
correlated genes identified by Pyro -Velocity, endocrine Beta cells
expressed insulin 1 and 2 (Ins1, Ins2) and Nnat, which is a key
regulator Beta cell insulin generation \citep{Millership2018-vd} and
Ppp1r1a which is a marker of Beta cells \citep{Jiang2013-pt}. The
ranking procedure proposed by scVelo failed to identify Ins1, Ins2,
Nnat, and Ppp1r1a as lineage-restricted genes in their top 50 ranked
gene list. The inability of scVelo to identify these key developmental
genes could be likely explained by the kNN smoothing of read count and
filtering steps (Supplementary Table 1, 2). Visual inspection of these
four top genes by Pyro -Velocity revealed a single activation phase
during cell type transitions based on phase portrait curves. We also
propose their visualization through ``rainbow plots'\,' to illustrate
denoised single gene spliced expression across shared time and to
investigate the inferred cell order based on known annotations. Further,
UMAP embedding of cells colored based on denoised gene expression
confirmed the cell fate markers' smooth and progressive activation (or
repression) towards endocrine cells (Fig.~\ref{fig-model-results}e,
Extended Data Fig.~\ref{extfig-model-results}).

@@

@@

\caption{\label{fig-model}Pyro -Velocity is a fully generative Bayesian
method with uncertainty estimation of velocity vector fields and shared
latent time, based on raw counts and without ad-hoc preprocessing steps.
a. Schematic comparing current RNA Velocity and Pyro -Velocity
workflows. Both methods take in input spliced and unspliced read count
tables and output RNA velocity vector fields and latent (shared) time.
Previous methods require ad-hoc preprocessing steps, including the
arbitrary selection of top variable genes, number of principal
components, and kNN-pooling to impute both spliced and unspliced
expression within the PCA-reduced spliced expression space. In addition,
manually tuned ad-hoc postprocessing steps are used to compute the final
vector field and latent time based on aggregation and smoothing of the
independent velocity estimates for each gene which can dramatically
change the velocity estimates. In contrast, Pyro -Velocity is a fully
generative and probabilistic model that uses raw read counts. Our model
requires minimal preprocessing and postprocessing steps and outputs a
common shared time based on all the genes without additional aggregation
steps. It also provides uncertainty estimation based on posterior
samples. b. The main outputs of the Pyro -Velocity framework.
Visualization of single gene expression across inferred developmental
cell order using rainbow plots (right panel).}

@@

\caption{\label{fig-model-results}Pyro -Velocity outperforms other
RNA-velocity methods and accurately predicts cell lineage trajectories.
a,b. RNA Velocity model comparisons from fully mature peripheral blood
mononuclear cells (PBMCs, t-SNE embedded plots, a) or E15.5 mouse
pancreas (UMAP embedded plots, b-c). From left to right in panels a and
b: 1. Cells colored by cell type; 2. Stream plot of velocity vector
field defined by scVelo; 3. Stream plot of the posterior mean vector
field from Pyro -Velocity; 4. Pyro -Velocity visualization of cell
dynamics for 3 arbitrarily selected single cells; 5. Single-cell vector
field uncertainty where each cell is colored based on the scaled angular
standard deviation to visualize uncertainty based on n=30 iterated,
posterior samples depicted as arrows noted throughout our analysis; 6.
Averaged vector field uncertainty where each vector is colored by the
averaged vector field uncertainty. The scaled angular standard deviation
range is {[}0 360{]}, where 0 corresponds to the highest confidence for
a single direction, 360 corresponds to the highest uncertainty, and no
directionality preference based on 30 posterior samples (4-6). Vectors
were projected into the UMAP space using the scVelo API based on cosine
similarity-derived transition matrix. c.~UMAP embedded cells colored by
the average cell shared time across 30 posterior samples. The Spearman's
correlation between Pyro -Velocity and Cytotrace is provided. d.
Pyro -Velocity automatic gene selection and ranking for predicting
important genes involved in pancreas development. e. Phase portraits of
posterior samples average spliced and unspliced gene expression (left),
rainbow plots showing denoised spliced gene expression (middle), and
gene expression within UMAP embedded space. Single dots correspond to
individual cells (left, right) or vertical lines (middle). Cell type
annotation coloring is the same as in panel b.}

@@

\caption{\label{fig-lineage-tracing}Pyro -Velocity accurately predicts
cell fate choices and lineage trajectory in human LARRY barcoded
hematopoietic cells. Systematic benchmarking of cell fate prediction
methods - predicting uni-lineage cell fates (a-b, Model 1), bifurcated
lineages (c, Model 2), or all lineages together (d, Model 2). Panels
from left to right show on the same SPRING embedding: 1. Cell types; 2.
Clonal progression vector field derived based on the observed clonal
barcodes (Methods); 3. scVelo recovered vector field; 4. Pyro -Velocity
single cell level vector field uncertainty (angular standard deviation
from 30 posterior samples); 5. Pyro -Velocity recovered vector field;
vectors are colored by the averaged vector field uncertainty; 6. Clonal
fate potency score using Cospar \citep{Wang2022-xb}; 7. scVelo recovered
latent time; 8. Pyro -Velocity inference of posterior mean shared time
per cell. Cosine similarity and Spearman's correlation for the different
methods with the clonal progression and fate potency are reported in the
titles.}

@@

@@

\caption{\label{extfig-graphical-models}Graphical representations detail
the generative process of the Pyro -Velocity models and highlight their
differences. a. Pyro -Velocity Model 1. Each rectangle represents one
data dimension (e.g., gene or cell). The arrows represent the directed
hierarchical dependency among random variables. The white circles
correspond to the modeled random variables. The probability
distributions of these variables are denoted on the right legend. The
gray circles correspond to intermediate results computed from random
variables. In this model, the kinetics random variables for the cell
shared time and gene are modeled with two constraints: 1. the shared
time is strictly larger than the gene-specific activation time
\(t_{0}^{(0)}\), 2. the gene expression in the activate state starts
with the initial condition \((u_{0}=0,\,s_{0}=0)\), as in previous
methods\citep{La_Manno2018-lj, Bergen2020-pj} (Methods). b.
Pyro -Velocity Model 2. This figure follows the same graphical
convention as in a. In this model \(t_{0}^{(0)}\) is independent from
the shared time, thus allowing a time lag for the gene activation phase.
In addition, a positive starting basal gene expression level
\((u_{0},\,s_{0})\) different than \((u_{0}=0,\,s_{0}=0)\) is allowed.
(Methods). c.~Schematic illustrating how the constraints proposed in
Model 1 (left) and Model 2 (right) are reflected in the modeling of
individual gene expression. The x-axis corresponds to the shared time,
the y-axis corresponds to the posterior mean of the spliced expression
level, the vertical lines with arrows show the start and end time of the
observed cell sampling, t0 represents the gene-specific activation time
\(t_{0}^{(0)}\), the red lines highlight the differences for the
potential gene-specific time lag when \(t_{0}^{(0)} > {\min}\!⁡(t)\), and
the initial spliced gene expression. The same differences can also be
illustrated for the unspliced gene expression profiles (not shown).}

@@

\caption{\label{extfig-model-results}Extended version of
Fig.~\ref{fig-model-results} for evaluating Pyro -Velocity (Model 1) and
scVelo on the PBMC and pancreas datasets. a. Left figure shows the
scVelo vector field using the default user options, right figure shows
the evaluation of the scVelo vector field using different combinations
of user options within Epsilon cells. b. Pyro -Velocity additional
single cell level uncertainty diagnostic plots on the fully mature PBMC
dataset, from left to right: 1. standard deviation of posterior samples
of shared time, 2. Rayleigh test of posterior samples of the vector
field (the title reports false discovery rate after p-value correction
for multiple tests across cells), 3. transcriptional state uncertainty
of predicted raw read count per cell. The contour lines show the top
10\% of the most uncertain cells. c.~For the pancreas dataset, the
figures show from left to right: UMAP rendering of 1. Cytotrace, 2.
scVelo latent time, 3. Posterior mean of Pyro -Velocity shared time
averaged across 30 posterior samples, 4. Top 50 markers overlapping
between Pyro -Velocity and scVelo (Methods, Supplementary Table 1-2), 5.
The same uncertainty diagnostic plots per cell as in b. d.~The same plot
as Fig. 2d. e. Selection of the top negatively correlated genes with the
cell shared time based on the posterior mean on denoised spliced
expression and negative mean absolute error of gene expression
prediction (Methods). f.~Phase portraits, rainbow plots, and UMAP
rendering of posterior mean of gene expression for the genes selected in
e.}

@@

\caption{\label{extfig-lineage-tracing}Extension of
Fig.~\ref{fig-lineage-tracing}, Pyro -Velocity (Model 1 for uni-fate
data, Model 2 for bi-fate and multi-fate data) outperforms Cytotrace on
scRNA-seq datasets with lineage barcoding information (LARRY dataset).
a. SPRING embedding rendering of Cytotrace scores, the title reports the
Spearman's correlation between the Cytotrace and Cospar fate potency
scores, b. Pyro -Velocity additional single cell level uncertainty
diagnostic plots based on the multi-fate LARRY dataset, from left to
right: 1. Standard deviation of posterior samples of shared time per
cell, 2. Rayleigh test of posterior samples of vector field per cell
(the title reports the false discovery rate after p-value correction for
multiple tests across cells), 3. Transcriptional state uncertainty of
predicted raw read count per cell. The contour lines show the top 10\%
of the most uncertain cells (Methods), c.~Pairwise Spearman's
correlation between Cytotrace/Cospar fate potency scores and velocity
methods for the four different benchmarking datasets selected based on
the lineage barcoding information from LARRY data. d.~SPRING embedding
rendering of posterior mean of shared time across 30 posterior samples,
the title shows the Spearman's correlation of mean shared time with
Cytotrace. e. Selection of the top positively correlated genes with the
cell shared time based on the posterior mean of spliced expression.
f.~Phase portraits, rainbow plots, and UMAP rendering with splicing gene
expression level for the selected genes in e.}

@@

\caption{\label{extfig-lineage-tracing-2}Extension of
Fig.~\ref{fig-lineage-tracing} presenting an extended comparison of
Pyro -Velocity and scVelo on scRNA-seq datasets with lineage barcoding
information (LARRY dataset). a. uni-fate Monocyte lineage analysis with
Pyro -Velocity Model 2, b. uni-fate Neutrophil lineage analysis with
Pyro -Velocity Model 2, c.~Neutrophil and Monocyte bi-fate lineage
analysis with Pyro -Velocity Model 1, and d.~multi-fate lineage analysis
with Pyro -Velocity Model 1. The panel descriptions are the same as in
Fig.~\ref{fig-lineage-tracing}.}

>>>>

\usepackage{graphicx}%
\usepackage{multirow}%
\usepackage{amsmath,amssymb,amsfonts}%
\usepackage{amsthm}%
\usepackage{mathrsfs}%
\usepackage[title]{appendix}%
\usepackage[dvipsnames]{xcolor}%
\usepackage{textcomp}%
\usepackage{manyfoot}%
\usepackage{booktabs}%
\usepackage{algorithm}%
\usepackage{algorithmicx}%
\usepackage{algpseudocode}%
\usepackage{listings}%

@@

\raggedbottom
\usepackage{nameref}
\usepackage{placeins}
\usepackage[labelfont=bf]{caption}
% \usepackage{showframe}
% \usepackage{layout}
\allowdisplaybreaks
\makeatletter
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{plain}
\@ifundefined{c@chapter}{\newfloat{suppfig}{h}{loextfig}}{\newfloat{suppfig}{h}{loextfig}[chapter]}
\floatname{suppfig}{Extended Data Fig.}
\newcommand*\listofsuppfigs{\listof{suppfig}{List of Extended Data Figures}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

@@

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Pyro -Velocity: Probabilistic RNA Velocity inference from single-cell data},
  pdfkeywords={RNA velocity, single-cell RNA sequencing, probabilistic
generative modeling, cell fate, developmental trajectory inference},
  colorlinks=true,
  linkcolor={NavyBlue},
  filecolor={Maroon},
  citecolor={NavyBlue},
  urlcolor={NavyBlue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{document}
\maketitle
% \clearpage
% \layout
\newpage{}

\renewcommand*\contentsname{Contents}
{
\hypersetup{linkcolor=NavyBlue}
\setcounter{tocdepth}{2}
\tableofcontents
}

@@

We next used Pyro -Velocity to unbiasedly uncover potential cell fate
determinant markers genes within the E15.5 mouse pancreas. To this end,
we first selected the top 300 genes with the best goodness of fit
(negative mean absolute errors across cells). Next, we ranked them based
on the positive (or negative) correlation between inferred shared time
and corresponding denoised spliced expression per cell
(Fig.~\ref{fig-model-results}d, \hyperref[sec-methods]{Methods}). This
procedure successfully recovered key developmental genes that are
activated (or repressed) during endocrinogenesis
(Fig.~\ref{fig-model-results}d-e,
Extended Data Fig.~\ref{extfig-model-results}). Among the top positively
correlated genes identified by Pyro -Velocity, endocrine Beta cells
expressed insulin 1 and 2 (Ins1, Ins2) and Nnat, which is a key
regulator Beta cell insulin generation \citep{Millership2018-vd} and
Ppp1r1a which is a marker of Beta cells \citep{Jiang2013-pt}. The
ranking procedure proposed by scVelo failed to identify Ins1, Ins2,
Nnat, and Ppp1r1a as lineage-restricted genes in their top 50 ranked
gene list. The inability of scVelo to identify these key developmental
genes could be likely explained by the kNN smoothing of read count and
filtering steps (Supplementary Table 1, 2). Visual inspection of these
four top genes by Pyro -Velocity revealed a single activation phase
during cell type transitions based on phase portrait curves. We also
propose their visualization through ``rainbow plots'\,' to illustrate
denoised single gene spliced expression across shared time and to
investigate the inferred cell order based on known annotations. Further,
UMAP embedding of cells colored based on denoised gene expression
confirmed the cell fate markers' smooth and progressive activation (or
repression) towards endocrine cells (Fig.~\ref{fig-model-results}e,
Extended Data Fig.~\ref{extfig-model-results}).

@@

\includegraphics{figures/v2/Figure1.pdf}

@@

\caption{\label{fig-model}Pyro -Velocity is a fully generative Bayesian
method with uncertainty estimation of velocity vector fields and shared
latent time, based on raw counts and without ad-hoc preprocessing steps.
\textbf{a}, Schematic comparing current RNA Velocity and Pyro -Velocity
workflows. Both methods take in input spliced and unspliced read count
tables and output RNA velocity vector fields and latent (shared) time.
Previous methods require ad-hoc preprocessing steps, including the
arbitrary selection of top variable genes, number of principal
components, and kNN-pooling to impute both spliced and unspliced
expression within the PCA-reduced spliced expression space. In addition,
manually tuned ad-hoc postprocessing steps are used to compute the final
vector field and latent time based on aggregation and smoothing of the
independent velocity estimates for each gene which can dramatically
change the velocity estimates. In contrast, Pyro -Velocity is a fully
generative and probabilistic model that uses raw read counts. Our model
requires minimal preprocessing and postprocessing steps and outputs a
common shared time based on all the genes without additional aggregation
steps. It also provides uncertainty estimation based on posterior
samples. \textbf{b}, The main outputs of the Pyro -Velocity framework.
Visualization of single gene expression across inferred developmental
cell order using rainbow plots (right panel).}

@@

\caption{\label{fig-model-results}Pyro -Velocity outperforms other
RNA-velocity methods and accurately predicts cell lineage trajectories.
\textbf{a,b}, RNA Velocity model comparisons from fully mature
peripheral blood mononuclear cells (PBMCs, t-SNE embedded plots, a) or
E15.5 mouse pancreas (UMAP embedded plots, \textbf{b-c}). From left to
right in panels \textbf{a} and \textbf{b}: 1. Cells colored by cell
type; 2. Stream plot of velocity vector field defined by scVelo; 3.
Stream plot of the posterior mean vector field from Pyro -Velocity; 4.
Pyro -Velocity visualization of cell dynamics for 3 arbitrarily selected
single cells; 5. Single-cell vector field uncertainty where each cell is
colored based on the scaled angular standard deviation to visualize
uncertainty based on n=30 iterated, posterior samples depicted as arrows
noted throughout our analysis; 6. Averaged vector field uncertainty
where each vector is colored by the averaged vector field uncertainty.
The scaled angular standard deviation range is {[}0 360{]}, where 0
corresponds to the highest confidence for a single direction, 360
corresponds to the highest uncertainty, and no directionality preference
based on 30 posterior samples (4-6). Vectors were projected into the
UMAP space using the scVelo API based on cosine similarity-derived
transition matrix. \textbf{c}, UMAP embedded cells colored by the
average cell shared time across 30 posterior samples. The Spearman's
correlation between Pyro -Velocity and Cytotrace is provided.
\textbf{d}, Pyro -Velocity automatic gene selection and ranking for
predicting important genes involved in pancreas development. \textbf{e},
Phase portraits of posterior samples average spliced and unspliced gene
expression (left), rainbow plots showing denoised spliced gene
expression (middle), and gene expression within UMAP embedded space.
Single dots correspond to individual cells (left, right) or vertical
lines (middle). Cell type annotation coloring is the same as in
\textbf{b}.}

@@

\caption{\label{fig-lineage-tracing}Pyro -Velocity accurately predicts
cell fate choices and lineage trajectory in human LARRY barcoded
hematopoietic cells. Systematic benchmarking of cell fate prediction
methods - predicting uni-lineage cell fates (\textbf{a-b}, Model 1),
bifurcated lineages (\textbf{c}, Model 2), or all lineages together
(\textbf{d}, Model 2). Panels from left to right show on the same SPRING
embedding: 1. Cell types; 2. Clonal progression vector field derived
based on the observed clonal barcodes (\hyperref[sec-methods]{Methods});
3. scVelo recovered vector field; 4. Pyro -Velocity single cell level
vector field uncertainty (angular standard deviation from 30 posterior
samples); 5. Pyro -Velocity recovered vector field; vectors are colored
by the averaged vector field uncertainty; 6. Clonal fate potency score
using Cospar \citep{Wang2022-xb}; 7. scVelo recovered latent time; 8.
Pyro -Velocity inference of posterior mean shared time per cell. Cosine
similarity and Spearman's correlation for the different methods with the
clonal progression and fate potency are reported in the titles.}

@@

\includegraphics{figures/v2/Supplementary-Figure1.pdf}

@@

\caption{\label{extfig-graphical-models}Graphical representations detail
the generative process of the Pyro -Velocity models and highlight their
differences. \textbf{a}, Pyro -Velocity Model 1. Each rectangle
represents one data dimension (e.g., gene or cell). The arrows represent
the directed hierarchical dependency among random variables. The white
circles correspond to the modeled random variables. The probability
distributions of these variables are denoted on the right legend. The
gray circles correspond to intermediate results computed from random
variables. In this model, the kinetics random variables for the cell
shared time and gene are modeled with two constraints: 1. the shared
time is strictly larger than the gene-specific activation time
\(t_{0}^{(0)}\), 2. the gene expression in the activate state starts
with the initial condition \((u_{0}=0,\,s_{0}=0)\), as in previous
methods\citep{La_Manno2018-lj, Bergen2020-pj}
(\hyperref[sec-methods]{Methods}). \textbf{b}, Pyro -Velocity Model 2.
This figure follows the same graphical convention as in \textbf{a}, In
this model \(t_{0}^{(0)}\) is independent from the shared time, thus
allowing a time lag for the gene activation phase. In addition, a
positive starting basal gene expression level \((u_{0},\,s_{0})\)
different than \((u_{0}=0,\,s_{0}=0)\) is allowed.
(\hyperref[sec-methods]{Methods}). \textbf{c}, Schematic illustrating
how the constraints proposed in Model 1 (left) and Model 2 (right) are
reflected in the modeling of individual gene expression. The x-axis
corresponds to the shared time, the y-axis corresponds to the posterior
mean of the spliced expression level, the vertical lines with arrows
show the start and end time of the observed cell sampling, t0 represents
the gene-specific activation time \(t_{0}^{(0)}\), the red lines
highlight the differences for the potential gene-specific time lag when
\(t_{0}^{(0)} > {\min}\!⁡(t)\), and the initial spliced gene expression.
The same differences can also be illustrated for the unspliced gene
expression profiles (not shown).}

@@

\caption{\label{extfig-model-results}Extended version of
Fig.~\ref{fig-model-results} for evaluating Pyro -Velocity (Model 1) and
scVelo on the PBMC and pancreas datasets. \textbf{a}, Left figure shows
the scVelo vector field using the default user options, right figure
shows the evaluation of the scVelo vector field using different
combinations of user options within Epsilon cells. \textbf{b},
Pyro -Velocity additional single cell level uncertainty diagnostic plots
on the fully mature PBMC dataset, from left to right: 1. standard
deviation of posterior samples of shared time, 2. Rayleigh test of
posterior samples of the vector field (the title reports false discovery
rate after p-value correction for multiple tests across cells), 3.
transcriptional state uncertainty of predicted raw read count per cell.
The contour lines show the top 10\% of the most uncertain cells.
\textbf{c}, For the pancreas dataset, the figures show from left to
right: UMAP rendering of 1. Cytotrace, 2. scVelo latent time, 3.
Posterior mean of Pyro -Velocity shared time averaged across 30
posterior samples, 4. Top 50 markers overlapping between Pyro -Velocity
and scVelo (\hyperref[sec-methods]{Methods}, Supplementary Table 1-2),
5. The same uncertainty diagnostic plots per cell as in \textbf{b}.
\textbf{d}, The same plot as Fig.~\ref{fig-model-results}d, e. Selection
of the top negatively correlated genes with the cell shared time based
on the posterior mean on denoised spliced expression and negative mean
absolute error of gene expression prediction
(\hyperref[sec-methods]{Methods}). \textbf{f}, Phase portraits, rainbow
plots, and UMAP rendering of posterior mean of gene expression for the
genes selected in \textbf{e}.}

@@

\caption{\label{extfig-lineage-tracing}Extension of
Fig.~\ref{fig-lineage-tracing}, Pyro -Velocity (Model 1 for uni-fate
data, Model 2 for bi-fate and multi-fate data) outperforms Cytotrace on
scRNA-seq datasets with lineage barcoding information (LARRY dataset).
\textbf{a}, SPRING embedding rendering of Cytotrace scores, the title
reports the Spearman's correlation between the Cytotrace and Cospar fate
potency scores, \textbf{b}, Pyro -Velocity additional single cell level
uncertainty diagnostic plots based on the multi-fate LARRY dataset, from
left to right: 1. Standard deviation of posterior samples of shared time
per cell, 2. Rayleigh test of posterior samples of vector field per cell
(the title reports the false discovery rate after p-value correction for
multiple tests across cells), 3. Transcriptional state uncertainty of
predicted raw read count per cell. The contour lines show the top 10\%
of the most uncertain cells (\hyperref[sec-methods]{Methods}),
\textbf{c}, Pairwise Spearman's correlation between Cytotrace/Cospar
fate potency scores and velocity methods for the four different
benchmarking datasets selected based on the lineage barcoding
information from LARRY data, \textbf{d}, SPRING embedding rendering of
posterior mean of shared time across 30 posterior samples, the title
shows the Spearman's correlation of mean shared time with Cytotrace.
\textbf{e}, Selection of the top positively correlated genes with the
cell shared time based on the posterior mean of spliced expression.
\textbf{f}, Phase portraits, rainbow plots, and UMAP rendering with
splicing gene expression level for the selected genes in \textbf{e}.}

@@

\caption{\label{extfig-lineage-tracing-2}Extension of
Fig.~\ref{fig-lineage-tracing} presenting an extended comparison of
Pyro -Velocity and scVelo on scRNA-seq datasets with lineage barcoding
information (LARRY dataset). \textbf{a}, uni-fate Monocyte lineage
analysis with Pyro -Velocity Model 2, \textbf{b}, uni-fate Neutrophil
lineage analysis with Pyro -Velocity Model 2, \textbf{c}, Neutrophil and
Monocyte bi-fate lineage analysis with Pyro -Velocity Model 1, and
\textbf{d}, multi-fate lineage analysis with Pyro -Velocity Model 1. The
panel descriptions are the same as in Fig.~\ref{fig-lineage-tracing}.}
